---
title: "Loading Difficult Data"
subtitle: "BODL R Training"
date: "2023-02-16"
output: 
  ioslides_presentation:
    css: ../custom.css
    fig_caption: yes
    logo: assets/bodl_logo_white_background.jpg
    widescreen: yes
---

```{r, loadPackages, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, include = TRUE, 
  warning = FALSE, message = FALSE, 
  out.width = 800, fig.align = "center"
)
```

# Loading Difficult Data

## Setup

If you've started a new session since last time:

```{r}
library(tidyverse)
```


## What if the data we have isn't nice?

- Missing values might be given a value (e.g. 9999, "NA")
- Column names might be missing
- File may have comments
- Irrelevant columns
- May be structural errors in the file (Different column numbers)
- White-space in cells

## What if the data we have isn't nice?

- Download the dataset `transport.csv` to your `data` folder
- Navigate to the file in your `Files` pane
- Click on the file and choose `View File`

**What problems do we face here?**

```{r, echo=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## What if the data we have isn't nice? 

**What problems do we face here?**

1. A comment in the first line
2. No column names
3. Missing data encoded as "-" in the 4^th^ line
4. A redundant column

## What if the data we have isn't nice? 

Let's try writing code for this instead of using the GUI

**This will fail**

```{r}
transport <- read_csv("data/transport.csv")
```

- `R` uses the first row to guess how many columns there are
- The comment is indicating 1 column

## Removing Comments

- We can tell `R` to ignore any lines beginning with `#`
- Set the comment argument using `comment = "#"`

```{r}
transport <- read_csv("data/transport.csv", comment = "#")
transport
```

Now `R` is guessing the correct number of columns $\implies$ the file will load

**What does all that red (or blue) stuff mean?**

## Data without column names

- `R` has assumed the first row contains column names
- We can tell `R` to ignore these using: `col_names = FALSE`

```{r}
transport <- read_csv(
  "data/transport.csv", 
  comment = "#", col_names = FALSE
)
transport
```

**What has `R` used for column names?** 

## Missing Data

**What impact has the missing data in `X5` had?**

- We can correctly assign missing data as `NA` (`na = "-"`)

```{r}
transport <- read_csv(
  "data/transport.csv", 
  comment = "#", col_names = FALSE, na = "-"
)
transport
```

## Ignoring Columns

```{r, echo=FALSE}
knitr::opts_chunk$set(eval = TRUE)
```

- We can also use the code `-` to skip a column
- We can leave `R` to guess any remaining columns using `?`
- Use the `col_types` argument: `col_types = "-?????"`

```{r, results='hide'}
transport <- read_csv(
  "data/transport.csv",
  comment = "#", col_names = FALSE, na = "-",
  col_types = "-?????"
)
transport
```

## Specifying Columns

- Or we can specify the exact type of data in each column
- Numeric columns can be specified as `n`
- Text or `character` columns can be specified as `c`

```{r}
transport <- read_csv(
  "data/transport.csv", 
  comment = "#", col_names = FALSE, na = "-", 
  col_types = "-ccnnc"
)
transport
```

## Specifying Columns

- Let's accidentally **get the final column wrong** (use `n` instead of `c`)

```{r,eval=FALSE}
transport <- read_csv(
  "data/transport.csv",
  comment = "#", col_names = FALSE, na = "-",
  col_types = "-ccnnn"
)
transport
```

**NB: No warning will be given if a numeric column contains non-numeric characters**


## Specifying Columns

Let's change that back to the correct code:

```{r}
transport <- read_csv(
  "data/transport.csv", 
  comment = "#", col_names = FALSE, na = "-", 
  col_types = "-ccnnc"
)
transport
```

## Setting Column Names

- My fix is to supply a `vector` of names

```{r}
myNames <- c(
  "X1", "gender", "name", "weight", "height", "method"
  )
transport <- read_csv(
  "data/transport.csv", 
  comment = "#", col_names = myNames, na = "-", 
  col_types = "-ccnnc"
)
transport
```


## The `c()` function

The most common function in `R` is `c()`

- This stands for `combine`
- Combines all values into a single `R` object, or `vector`
- If left empty, it is equivalent to `NULL`

```{r, results='markup'}
c()
```

## The `c()` function

- I have used this to create a vector of column names 
- Used the assignment operator `<-` to assign this vector

**What would happen if I gave too many or too few names?**

- We need to be careful here...

## Exporting Data

After we've edited a file, we might also wish to export it

```{r, eval=FALSE}
?write_csv
```

- This is a wrapper for `write_delim()`
- Can export `.csv`, `.txt`, `.tsv` etc.
- Individual `R` objects can be exported using `write_rds()`

## Exporting Data

The best way to export this is:

```{r}
write_csv(transport, "data/transport_clean.csv")
```


<div class="footer" style="text-align:center;width:25%">
[Home](http://uofabioinformaticshub.github.io/RAdelaide-July-2018)
</div>
